# Reproducible Research assignment 1

## Loading and preprocessing the data

### Load the data

```{r load}
setwd("E:/repdata_data_activity")

activity <- read.csv("activity.csv")

str(activity)
```

## What is mean total number of steps taken per day?

### Calculate the total number of steps taken per day

``` {r summing1}
library(dplyr)

activity %>%
  group_by(date) %>%
  summarise(
    total_steps_per_day = sum(steps, na.rm = TRUE)
  ) -> total_steps_each_day 

total_steps_each_day
```
 
### Histogram of the total number of steps taken each day

```{r hist1}
hist(total_steps_each_day[,2])
```

### mean and median of the total number of steps taken per day

```{r summary1}
total_steps_each_day %>%
  summarise(
    mean_steps_per_day = mean(total_steps_per_day, na.rm = TRUE)
  ) 

total_steps_each_day %>%
  summarise(
    median_steps_per_day = median(total_steps_per_day, na.rm = TRUE)
  ) 
```

mean steps taken per day: 9354
median steps taken per day: 10395

## What is the average daily activity pattern?

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r mean1}
activity %>%
  group_by(interval) %>%
  summarise(
    avg_steps_per_interval = mean(steps, na.rm = TRUE)
  ) -> avg_steps_per_interval

plot(avg_steps_per_interval,type="l")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Sort the average steps per interval in descending order and the top row of the table will have the maximum number of steps.

``` {r max1}
avg_steps_per_interval %>% 
  arrange(desc(avg_steps_per_interval)) -> avg_steps_per_interval_desc

avg_steps_per_interval_desc[1,]
```
Interval at 835 minutes has highest number of steps taken averaged across all days

## Imputing missing values

### Calculate and report the total number of missing values in the dataset

```{r calc_missing}
colSums(is.na(activity))
```

There are 2304 missing values in the dataset

### Impute missing values with mean steps for that interval

Firstly, create a merged table to include the average steps per interval information to the original table
``` {r merge_table}
merged <- inner_join(activity,avg_steps_per_interval)

head(merged)
```

Impute missing data with the average steps per interval

``` {r impute}
for(i in 1:nrow(activity)){
  if(is.na(activity[i,1] == TRUE)){
  activity[i, 1] <- merged[i,4] 
  }
}

head(activity)
```

### Summary of imputed data
Make a histogram of the total number of steps taken each day and 

```{r summing2}
activity %>%
  group_by(date) %>%
  summarise(
    total_steps_per_day = sum(steps, na.rm = TRUE)
  ) -> total_steps_each_day 

total_steps_each_day

hist(total_steps_each_day[,2])
```

Calculate and report the mean and median total number of steps taken per day. 

```{r summary2}
total_steps_each_day %>%
  summarise(
    mean_steps_per_day = mean(total_steps_per_day, na.rm = TRUE)
  ) 

total_steps_each_day %>%
  summarise(
    median_steps_per_day = median(total_steps_per_day, na.rm = TRUE)
  ) 
```

Do these values differ from the estimates from the first part of the assignment? 
Yes, the values are different.

What is the impact of imputing missing data on the estimates of the total daily number of steps?
The values are increased slightly.

## Are there differences in activity patterns between weekdays and weekends?
Indicating whether a given date is a weekday or weekend day.

Firstly, determine the day from the date. Then determine if the day is a weekday or weekend.

```{r indicate}
activity$day <- weekdays(as.Date(activity$date))
head(activity)
weekday_or_weekend <- vector(mode = "list", length = 17568)

for(i in 1:nrow(activity)){
  if(activity[i,4] == "Saturday"){
  weekday_or_weekend[[i]] <- "weekend" } else if (activity[i,4] == "Sunday"){
  weekday_or_weekend[[i]] <- "weekend" } else {
  weekday_or_weekend[[i]] <- "weekday"
  }
}

activity$weekday_or_weekend <- weekday_or_weekend

head(activity)
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

Splitting the table into a weekday only table and weekend only table for plotting.
``` {r split_table}
activity %>% 
filter(weekday_or_weekend == "weekday") -> activity_weekday

dim(activity_weekday)

activity %>% 
filter(weekday_or_weekend == "weekend") -> activity_weekend

dim(activity_weekend)
```

Calculating the average steps per interval across the days for only weekdays and only weekends. Then plot them seperately.

```{r calculate}
activity_weekday %>%
  group_by(interval) %>%
  summarise(
    avg_steps_per_interval = mean(steps, na.rm = TRUE)
  ) -> avg_steps_per_interval_weekday

plot(avg_steps_per_interval_weekday,type="l",main="weekday average steps per interval")

activity_weekend %>%
  group_by(interval) %>%
  summarise(
    avg_steps_per_interval = mean(steps, na.rm = TRUE)
  ) -> avg_steps_per_interval_weekend

plot(avg_steps_per_interval_weekend,type="l",main="weekend average steps per interval")
```
